---
title: "dsa2101_rmdcompiled"
author: "hui xuan"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction (1 paragraph)

Briefly introduce the data set. 
You can repeat some of the data description found on the TidyTuesday repository, paraphrasing on your own terms. 
Imagine that your reader has no prior knowledge about the data set.

## Descriptive Statistics (1-2 paragraphs, a few code blocks)

Mention any transformation or cleaning you have applied to the data. 
Present key descriptive statistics to give a taste of what the data are like. 
Resist the temptation to report descriptive statistics in bulk and be more selective in reporting only the most interesting/relevant statistics.

```{r billboard, echo=FALSE}
billboard <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/billboard.csv')

audio_features <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/audio_features.csv')
```

```{r stats, echo=FALSE}
library(tidyverse)
library(stringr)

# shape, column types, summary, structure of the dataset
dim(billboard) # 10 columns of dataset with 327895 observations
str(billboard) # url, week_id, song, performer, song_id chr, the rest numerics
summary(select_if(billboard, is.numeric))

```

```{r transformations, echo=FALSE}
#clean and merge dataset
billboard1 <- billboard %>%
  drop_na()

audio_features1 <- audio_features %>%
  drop_na()

merged <- billboard1 %>%
  left_join(audio_features, by = "song_id", suffix = c("_board", "_features")) #perform a left join?
  
merged$week_id <- as.POSIXct(merged$week_id, format = "%d/%m/%Y") #converts the string to datetime woah
merged$week_id <- format(merged$week_id, format = "%Y")
```

## Question 1
### insert question here
#### introduction to the question

#### Methodology

##### 1st plot: line chart

(insert description of variables plotted, design choices for why the choice of variables and information used is the best for providing the answers to the question)

##### 2nd plot: bar chart

(insert description of variables plotted, design choices for why the choice of variables and information used is the best for providing the answers to the question)

#### Visualizations & Discussion

```{r q1a, echo=FALSE}
#insert code here
```

this plot means...

```{r q1b, echo=FALSE}
#insert code here
```

this plot means...

## Question 2
### insert question here
#### introduction to the question

#### Methodology

##### 1st plot: line chart

(insert description of variables plotted, design choices for why the choice of variables and information used is the best for providing the answers to the question)

##### 2nd plot: bar chart

(insert description of variables plotted, design choices for why the choice of variables and information used is the best for providing the answers to the question)

#### Visualizations & Discussion

```{r q2a, echo=FALSE}
#insert code here
```

this plot means...

```{r q2b, echo=FALSE}
#insert code here
```

this plot means...

## References
- reference the dataset
- other information sources should be added here
